Promise and peril: Agnostic identification methods\\ for detecting differential item functioning

### 2020-03-19 paper/paper_2b_anchor_points.Rmd
getting there

The previously discussed AI methods select a set of anchor items, whether it is an algorithm or the analyst that makes that selection. The anchor items are used to estimate $\hat \mu^\text{foc}$. An alternative strategy is to directly set the anchor point, $\mu^{\star\text{foc}}$. Anchor point methods have the advantage of not requiring the assumption that any particular item is DIF-free, and, therefore, allowing all items to be tested for DIF. The question then becomes the following: How is the anchor point selected?

@@

$\mu^{\star\text{foc}}$ is selected by maximizing the Gini index (thus the abbreviation MAXGI). The intuition and assumption is that the items without DIF are the most homogeneous. Denoting a function that calculates the Gini index from a vector of non-negative elements as  $G(\mathbf{x})$, MGI sets <!-- its weird to arg max over a parameter we already have a definition for - maybe switch to t or something -->
$$
\mu^{\star\text{foc}} = \argmax_{\mu^\text{foc}} G(|\mu^\text{foc} + \tilde{\mathbf{d}}|)
$$
where $\mu^\text{foc} \in (-\infty, \infty)$ and $\mu^\text{foc}$ is added to each element of $\mathbf{d}$.

@@

For our one-run simulation, Figure \ref{fig:ginipath} shows the gini coefficient at a variety of possible $\mu^\text{foc}$ values. The result of MAXGI is $\mu^{\star\text{foc}} = `r round(gini_results$gini_anchor_points, 2)`$, which is extraordinarily close to the data-generating value of -1.

@@

Raju's area method [@raju1988area] measures the amount of DIF by calculating the area between the item characteristic curves, the function that maps the student's ability to their probability of correct response, of the two groups:
$$
\text{Area Between Curves} = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, b_j^{\text{foc}})|
$$
Raju's area method has been cited as one of the most commonly used IRT-based DIF detection methods [@magis2011generalized]. However, Raju's area method is not an AI method because the item characteristic curves still need to be linked by anchor items or an anchor point. An additional weakness is that the area is unweighted, so all values of $\theta$ matter equally, despite some being much more realistic than others.

<!-- again might want to choose a different variable for provisional values -->
Denote a function that takes $\mu^\text{foc}$ as input and estimates $\hat b_j^\text{foc}$ by fitting a unidimensional Rasch model as $m_j(\mu^\text{foc})$. The amount of DIF on each item is calculated as
$$
\text{DIF}_j(\mu^\text{foc}) = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, m_j(\mu^\text{foc}))| g(\theta)d\theta
$$
where $g(\theta)$ is a weighting function such that $\int g(\theta)d\theta = 1$. The total DIF on the test, then, is
$$
\text{Total DIF}(\mu^\text{foc}) = \sum_{j} \text{DIF}_j(\mu^\text{foc})
$$
In this way, $\text{Total DIF}(\mu^\text{foc})$ is a function where the input is $\mu^\text{foc}$ and the output is the total amount of DIF on the test. MINBC sets
$$
\mu^{\star\text{foc}} = \argmin_{\mu^\text{foc}} \text{Total DIF}(\mu^\text{foc}).
$$
MINBC is inspired in part by @chalmers2016might, who use the difference between test characteristic curves weighted by $g(\theta)$ as a measure of differential test functioning (DTF). The selection of $g(\theta)$ results in the relative weighting of $\theta$ values. Chalmers, Counsell, and Flora do not discuss how to choose $g(\theta)$ and in their empirical examples use $g(\theta)$ uniform for $-6 \le \theta \le 6$, which may be suboptimal in some cases. It might seem intuitive to choose $g(\theta) \sim N(0, 1)$ because $\mu^\text{ref} = 0$, but this choice doesn't take into account the ability distribution of the focal group. If $\mu^\text{foc} = 3$, wouldn't we also want to prioritize high $\theta$ values? Accordingly, we set $g(\theta)$ to be the average of the reference and focal group ability probability density functions:
$$
g(\theta) = \dfrac{N(\mu^{\text{ref}}, \sigma^{\text{ref}^2}) + N(\mu^{\text{foc}}, \sigma^{\text{foc}^2})}{2}.
$$

@@

For our one-run simulation, Figure \ref{fig:mabc} shows Total DIF at a variety of possible values for $\mu^\text{foc}$. In this case, MINBC works perfectly and the anchor point is found to be $\mu^{\star\text{foc}} = `r round(min_bc$between_curves_anchor_points, 2)`$. As with MAXGI, the model should then be refit using the identifying assumption that $\mu^{\text{foc}} = \mu^{\star\text{foc}}$.

@@

equal means, multiple imputation logit graph (EM-MILG) & Arbitrarily set both group means to 0, which pushes all group performance differences to the item paramaters, measure variability using multiple imputations, and graph the result. Can be used by an analyst to hand select anchor items & Inspired by pedagolical examples \cite{pohl2017cluster} and \cite{talbot2013taking} \\\hline

>>>>

The previously discussed AI methods select a set of anchor items, whether it is an algorithm or the analyst that makes that selection. The anchor items are used to estimate $\hat \mu^\text{foc}$. An alternative strategy is to directly set the anchor point [@strobl2018anchor]. Anchor point methods have the advantage of not requiring the assumption that any particular item is DIF-free, and, therefore, allowing all items to be tested for DIF. The question then becomes the following: How is the anchor point selected?

@@

$\mu^{\star\text{foc}}$ is selected by maximizing the Gini index (thus the abbreviation MAXGI). The intuition and assumption is that the items without DIF are the most homogeneous. Denoting a function that calculates the Gini index from a vector of non-negative elements as  $G(\mathbf{x})$, MGI sets <!-- its weird to arg max over a parameter we already have a definition for - maybe switch to t or something -->
\begin{align}
\mu^{\star\text{foc}} = \argmax_{\mu^\text{foc}} G(|\mu^\text{foc} + \tilde{\mathbf{d}}|)
\end{align}
where $\mu^\text{foc} \in (-\infty, \infty)$ and $\mu^\text{foc}$ is added to each element of $\mathbf{d}$.

@@

For our demonstration data simulation, Figure \ref{fig:ginipath} shows the gini coefficient at a variety of possible $\mu^\text{foc}$ values. The result of MAXGI is $\mu^{\star\text{foc}} = `r round(gini_results$gini_anchor_points, 2)`$, which is extraordinarily close to the data-generating value of -1.

@@

@@

Raju's area method [@raju1988area] measures the amount of DIF by calculating the area between the item characteristic curves, the function that maps the student's ability to their probability of correct response, of the two groups:
\begin{align}
\text{Area Between Curves} = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, b_j^{\text{foc}})|
\end{align}
Raju's area method has been cited as one of the most commonly used IRT-based DIF detection methods [@magis2011generalized]. However, Raju's area method is not an AI method because the item characteristic curves still need to be linked by anchor items or an anchor point. An additional weakness is that the area is unweighted, so all values of $\theta$ matter equally, despite some being much more realistic than others.

<!-- again might want to choose a different variable for provisional values -->
Denote a function that takes $\mu^\text{foc}$ as input and estimates $\hat b_j^\text{foc}$ by fitting a unidimensional Rasch model as $m_j(\mu^\text{foc})$. The amount of DIF on each item is calculated as
\begin{align}
\text{DIF}_j(\mu^\text{foc}) = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, m_j(\mu^\text{foc}))| g(\theta)d\theta
\end{align}
where $g(\theta)$ is a weighting function such that $\int g(\theta)d\theta = 1$. The total DIF on the test, then, is
\begin{align}
\text{Total DIF}(\mu^\text{foc}) = \sum_{j} \text{DIF}_j(\mu^\text{foc})
\end{align}
In this way, $\text{Total DIF}(\mu^\text{foc})$ is a function where the input is $\mu^\text{foc}$ and the output is the total amount of DIF on the test. MINBC sets
\begin{align}
\mu^{\star\text{foc}} = \argmin_{\mu^\text{foc}} \text{Total DIF}(\mu^\text{foc}).
\end{align}

MINBC is inspired in part by @chalmers2016might, who use the difference between test characteristic curves weighted by $g(\theta)$ as a measure of differential test functioning (DTF). The selection of $g(\theta)$ results in the relative weighting of $\theta$ values. Chalmers, Counsell, and Flora do not discuss how to choose $g(\theta)$ and in their empirical examples use $g(\theta)$ uniform for $-6 \le \theta \le 6$, which may be suboptimal in some cases. It might seem intuitive to choose $g(\theta) \sim N(0, 1)$ because $\mu^\text{ref} = 0$, but this choice doesn't take into account the ability distribution of the focal group. If $\mu^\text{foc} = 3$, wouldn't we also want to prioritize high $\theta$ values? Accordingly, we set $g(\theta)$ to be the average of the reference and focal group ability probability density functions:
\begin{align}
g(\theta) = \dfrac{N(\mu^{\text{ref}}, \sigma^{\text{ref}^2}) + N(\mu^{\text{foc}}, \sigma^{\text{foc}^2})}{2}.
\end{align}

For our demonstration data simulation, Figure \ref{fig:mabc} shows Total DIF at a variety of possible values for $\mu^\text{foc}$. In this case, MINBC works perfectly and the anchor point is found to be $\mu^{\star\text{foc}} = `r round(min_bc$between_curves_anchor_points, 2)`$. As with MAXGI, the model should then be refit using the identifying assumption that $\mu^{\text{foc}} = \mu^{\star\text{foc}}$.

@@

### 2020-03-19 paper/paper_3_sim.Rmd
getting there

In our simulation study, it was critical that student ability was drawn in a realistic way similar to @walker2017using. However, we don’t believe that a compensatory model is realistic in describing a math test where some items depend on verbal ability. For example, it’s hard to imagine that a student without the verbal ability to parse a word problem could fully compensate by having a higher math ability. Accordingly, we generate item responses using a simplified version of Sympson’s -@sympson1978model noncompensatory item response model in which
$$
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{2j}) = \sigma(\theta_i) \cdot \sigma(a_{2j}\eta_i)
$$
where, as before, $\theta_i$ is target ability, $\eta_i$ is nuisance ability, and $\sigma(a_{2j})$ is the item's loading on nuisance ability [@demars2016partially].

@@

In each run, we simulate 10,000 students with half coming from each of the reference and focal groups. For students from the reference group, target ability and nuisance ability are drawn from the two-dimensional normal distribution with mean [$\mu_\theta^\text{ref} = 0$, $\mu_\eta^\text{ref} = 0$] and covariance matrix $\begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix}$. Abilities for students from the focal group are drawn using the same covariance matrix, but with means [$\mu_\theta^\text{foc} = -0.5$, $\mu_\eta^\text{foc} = -1$].

The test always has 12 items, but we vary the number of items with DIF from two to six. For items without DIF, $a_{2j} = \infty$ so that the model reduces to $\text{Pr}(y_{ij} = 1 | \theta_i) = \sigma(\theta_i)$. For items with DIF, $a_{2j}$ is calculated based on Ackerman's -@ackerman1994using angle equation as described in @walker2017using:
$$
\angle_j = \arccos \dfrac{a_{1j}^2}{a_{1j}^2 + a_{2j}^2}.
$$
An item's angle measures the relative loading of the item on the two dimensions. For example, an angle of 45$^\circ$ indicates that the item loads equally on the target ability and nuisance ability. Our simple noncompensatory model has $a_{1j} = 1$ for all items so the angle equation reduces to
$$
\angle_j = \arccos \dfrac{1}{1 + a_{2j}^2}.
$$
We are interested in specifying the angle of an item, so the relevant equation becomes
$$
a_{2j} = \sqrt{\dfrac{1 - \cos(\angle_j)^2}{\cos(\angle_j)^2}}.
$$
For DIF items, we set $a_{2j}$ based on angles with equal intervals between 20$^\circ$ and 60$^\circ$. For example, for a test with three DIF items the angles are 20$^\circ$, 40$^\circ$, and 60$^\circ$.

@@

Figure \ref{fig:anchorfalse} shows the mean false anchor rates for each method and number of items with DIF. For example, when there were two items with DIF on the test, those two items had $\angle_{11} = 20^\circ$ and $\angle_{12} = 60^\circ$. AOAA-OAT never included item 12 in the anchor set, but incorrectly included item 11 in 60 out of the 100 runs. Accordingly, the mean false anchor rate for two DIF items and the AOAA-OAT method was
$$
\dfrac{\text{Total number of DIF items in the anchor set}}{\text{Number of DIF items on each test} \cdot \text{Number of runs}} = \dfrac{60}{2 \cdot 100} = 30\%.
$$
The fact that the item with 20$^\circ$ of DIF is most commonly incorrectly included in the anchor set is what drove the counterintuitive result that the mean false anchor rate decreases with more DIF items.

>>>>

In our simulation study, it was critical that student ability was drawn in a realistic way similar to @walker2017using. However, we don’t believe that a compensatory model is realistic in describing a math test where some items depend on verbal ability. For example, it’s hard to imagine that a student without the verbal ability to parse a word problem could fully compensate by having a higher math ability. Accordingly, we generated item responses using a simplified version of Sympson’s -@sympson1978model noncompensatory item response model in which
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{2j}) = \sigma(\theta_i) \cdot \sigma(a_{2j}\eta_i)
\end{align}
where, as before, $\theta_i$ is target ability, $\eta_i$ is nuisance ability, and $\sigma(a_{2j})$ is the item's loading on nuisance ability [@demars2016partially].

@@

In each run, we simulated 10,000 students with half coming from each of the reference and focal groups. For students from the reference group, target ability and nuisance ability are drawn from the two-dimensional normal distribution with mean [$\mu_\theta^\text{ref} = 0$, $\mu_\eta^\text{ref} = 0$] and covariance matrix $\begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix}$. Abilities for students from the focal group are drawn using the same covariance matrix, but with means [$\mu_\theta^\text{foc} = -0.5$, $\mu_\eta^\text{foc} = -1$].

The test always has 12 items, but we varied the number of items with DIF from two to six. For items without DIF, $a_{2j} = \infty$ so that the model reduces to $\text{Pr}(y_{ij} = 1 | \theta_i) = \sigma(\theta_i)$. For items with DIF, $a_{2j}$ is calculated based on Ackerman's -@ackerman1994using angle equation as described in @walker2017using:
\begin{align}
\angle_j = \arccos \dfrac{a_{1j}^2}{a_{1j}^2 + a_{2j}^2}.
\end{align}
An item's angle measures the relative loading of the item on the two dimensions. For example, an angle of 45$^\circ$ indicates that the item loads equally on the target ability and nuisance ability. Our simple noncompensatory model has $a_{1j} = 1$ for all items so the angle equation reduces to
\begin{align}
\angle_j = \arccos \dfrac{1}{1 + a_{2j}^2}.
\end{align}
We are interested in specifying the angle of an item, so the relevant equation becomes
\begin{align}
a_{2j} = \sqrt{\dfrac{1 - \cos(\angle_j)^2}{\cos(\angle_j)^2}}.
\end{align}
For DIF items, we set $a_{2j}$ based on angles with equal intervals between 20$^\circ$ and 60$^\circ$. For example, for a test with three DIF items the angles are 20$^\circ$, 40$^\circ$, and 60$^\circ$.

@@

Figure \ref{fig:anchorfalse} shows the mean false anchor rates for each method and number of items with DIF. For example, when there were two items with DIF on the test, those two items had $\angle_{11} = 20^\circ$ and $\angle_{12} = 60^\circ$. AOAA-OAT never included item 12 in the anchor set, but incorrectly included item 11 in 60 out of the 100 runs. Accordingly, the mean false anchor rate for two DIF items and the AOAA-OAT method was
\begin{align}
\dfrac{\text{Total number of DIF items in the anchor set}}{\text{Number of DIF items on each test} \cdot \text{Number of runs}} = \dfrac{60}{2 \cdot 100} = 30\%.
\end{align}
The fact that the item with 20$^\circ$ of DIF is most commonly incorrectly included in the anchor set is what drove the counterintuitive result that the mean false anchor rate decreases with more DIF items.
